Estimating the dynamics of\\SPARQL query results over Wikidata\thanks{This work was supported by CONICYT PFCHA, Doctorado Becas Chile/2017 -- 72180000, the Millennium Institute for Foundational Research on Data (IMFD) and FONDECYT Grant No. 1181896.}

### 2019-01-28 kgswc19.tex
Second quick pass on the paper

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{enumitem}
\usepackage{gensymb}
\usepackage{amsmath}
\usepgfplotslibrary{clickable}

@@

\begin{document}
%
\title{Estimating the dynamics of\\SPARQL query results over Wikidata\thanks{This work was supported by CONICYT PFCHA, Doctorado Becas Chile/2017 -- 72180000, the Millennium Institute for Foundational Research on Data (IMFD) and FONDECYT Grant No. 1181896.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Alberto Moya \and
Aidan Hogan\orcidID{0000-0001-9482-1982}}
%
\authorrunning{A. Moya and A. Hogan}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{IMFD; DCC, University of Chile\\
\email{\{amoya,ahogan\}@dcc.uhile.cl}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
In this work, we tackle the problem of estimating future changes in SPARQL query results over dynamic datasets; such work has applications for caching and remote synchronization. Our proposal is based on syntactic query features and (optionally) knowledge of the dynamics of predicates in the underlying dataset; the resulting model can be used to predict (1) whether or not the results for a query will change in the near future using classifiers, and (2) when they might change (if at all) using linear regression. To evaluate our proposal, we select Wikidata as our dataset and build a curated collection of real-world SPARQL queries. Our results show that the accuracy of the predictions of our proposed model is competitive with that obtained using knowledge of the complete historical evolution of the query results but at a lower cost.

@@

Some works \ah{Which? Cite at the end of the sentence} have proposed to predict dynamics based on the historical evolution of entities, but following such an approach for SPARQL queries is expensive because (1) a SPARQL query may involve potentially many entities; and (2) it is necessary to have the previous complete versions of data to analyse the entities relevant to a query. Recent works have explored the dynamics of Linked Data with the intention of finding patterns that allow for characterizing, recognizing, and predicting changes based on analysis of the domains, predicates, and schema~\cite{UmbrichHHPD10,KaferAUOH13,NishiokaS16,NishiokaS17,GonzalezH18}. Among these works, we can find hybrid approaches that have been developed to return fresher query results with faster response times by decomposing a query into dynamic sub-queries answered remotely, and static sub-queries answer over local caches~\cite{UmbrichKHP12}, but such an approach only focuses on estimating the dynamics of individual triple patterns, and does not consider, for example, query operators. More generally, coping with changes in remote data still presents a major challenge for applications leveraging dynamic Linked Data.

In this work, we tackle the problem of estimating future changes in query results based on query features, high-level statistics about the dynamics of predicates, as well as classification and regression methods from Machine Learning. We thus propose a solution where we do not have complete historical information about the results of a query, but rather a much more lightweight statistical model that saves on space and obviates the need to compute query results for multiple past versions. In order to assess the predictions made by our proposal, we perform experiments over 23 weekly versions of the Wikidata knowledge graph~\cite{VrandecicK14}, evaluating various configurations in terms of the accuracy for predicting future changes in the query results; we select Wikidata as: (1) it provides a history of weekly versions that we can use to evaluate predictions; (2) it is edited by thousands of users, which means that significant changes are observed weekly; (3) it provides a large-scale diverse dataset; (4) it receives millions of queries on a daily basis~\cite{MalyshevKGGB18}; and (5) a diverse set of exemplar SPARQL queries has been proposed by the community that we can use for our evaluation.

The contributions of this paper are thus as follows: (i) we propose a supervised method that uses static query features and suitable classifiers to predict whether or not results will change in a given point in the near future; (ii) we propose to augment this model with lightweight statistics of the dynamics of predicates; (iii) we further experiment with linear regression models to predict when results will change; (iv) we prepare a benchmark for predicting the dynamics of query results based on 23 weekly versions of Wikidata and a set of 221 real-world SPARQL queries; (v) we use this benchmark to compare the results obtained by our lightweight models versus an ``clairvoyant model''~\cite{KnuthHS16} that knows of changes in the query's results for previous versions. Our results show that our proposed models are competitive with the costly clairvoyant model.

The rest of this document is organized as follows: In Section~\ref{sec:related}, we review related works. Section~\ref{sec:preliminar} provides a formal definition of the problem. Section~\ref{sec:approach} describes our proposed models. In Section~\ref{sec:data}, we describe our evaluation data and queries based on Wikidata. Section~\ref{sec:eval} presents the design and results of the experiments. We conclude and review future directions in Section~\ref{sec:conclusion}.

@@

Specifically regarding the dynamics of SPARQL query results, Passant and Mendes~\cite{PassantM10} proposed \textsc{sparqlPuSH} as a notification framework based  on \textsc{PubSubHubBub} (recently standardized as \textsc{WebSub}~\cite{webSub18}) aiming for strong consistency. On the other hand, aiming for weak consistency, Umbrich et al.~\cite{UmbrichKHP12,UmbrichKPPH12,ekawUmbrichKHP12} proposed various methods to obtain knowledge about dynamics for different query patterns, mainly based on predicates. Dehghanzadeh et al.~\cite{DehghanzadehPKUHD14} later proposed a method to estimate the freshness using cardinality estimation techniques based on predicates and characteristic sets. Combining the notion of subscription-based notifications and predicting dynamics, Knuth et al.~\cite{KnuthHS16} propose a middleware to which consumers may subscribe and that periodically ranks and schedules refreshes for queries according to policies that take into account how likely the results are to be stale, how long ago the query was last refreshed, how many results previously changed, how long the query takes to run, etc.

Our proposal aims to model the dynamics of SPARQL query results in order to predict (1) whether or not the results will have changed in a fixed point in the near future and,  (2) when the results will change. Our work thus complements existing works aiming for weak consistency, but rather investigates the potential of using static features of a query as a lightweight alternative with the intuition that, for example, queries with more triple patterns or certain query operators are more likely to have dynamic results. For the purposes of comparison, we complement these static features with further knowledge of the dynamics of predicates~\cite{UmbrichKHP12,UmbrichKPPH12,ekawUmbrichKHP12,DehghanzadehPKUHD14} and knowledge of how results for the input query have changed over past versions~\cite{KnuthHS16}, as considered in previous works. We feed these data into Machine Learning classification and regression models to predict future changes (hence, for the purposes of supervised learning, we require knowledge of the changes of some queries over time, though not necessarily for the input query). The benefits of such an approach can be seen in a number of ways: (i) as complementing existing approaches; (ii) as an alternative to predicate-based statistics, which requires analysis of prior versions of the underlying data; (iii) as an alternative to historical results, which requires having a record of (or computing at runtime) the results of the query over several historical versions. We further remark that, to the best of our knowledge, no existing work aims to predict the dynamics of SPARQL query results using Machine Learning models.

@@

SPARQL is the recommended query language to retrieve and manipulate data stored in the RDF format. In this work, we focus on SPARQL \texttt{SELECT} queries, where we will first define a SPARQL 1.0 query. Let $\Vars$ be a set of variables disjoint from the set of RDF terms. A \emph{SPARQL expression} is built recursively as follows. (1) A \emph{triple pattern} $t \in  (\Iris \cup \Bnodes \cup \Lits \cup \Vars) \times (\Iris \cup \Vars) \times (\Iris \cup \Lits \cup \Bnodes \cup \Vars)$ is an \emph{expression}. (2) If $Q_1$ and $Q_2$ are expressions and $R$ is a filter condition, then $Q_1$ \texttt{FILTER} $R$, $Q_1$ \texttt{UNION} $Q_2$, $Q_1$ \texttt{OPTIONAL} $Q_2$, $Q_1$ \texttt{AND} $Q_2$ are $expressions$. Finally, if $Q$ is an expression, $V$ a list of variables and $\Delta$ a boolean value, $\texttt{SELECT}_V^\Delta Q$ is a SPARQL \texttt{SELECT} query, where $V$ denotes the projected variables, and $\Delta$ the \texttt{DISTINCT} option that when true, removes duplicate results. The semantics of a SPARQL \texttt{SELECT} query $Q$ is defined in terms of its \emph{evaluation} over an RDF graph $G$, denoted $Q(G)$, giving a set of partial mappings from projected variables to the set $\Iris \cup \Lits \cup \Bnodes$; we refer to P{\'{e}}rez et al.~\cite{PerezAG09} for definitions. (According to the standard, a SPARQL query is evaluated over a set of \emph{named graphs}; however, since the Wikidata query service indexes data in one RDF graph, to keep discussion concise, we consider RDF graphs, though our methods can be generalized in a straighforward manner to RDF datasets/named graphs.)

Our method supports SPARQL 1.1 \texttt{SELECT} queries, which allow a variety of additional features. One key feature in this extension is that of \textit{property paths}, which allows for matching arbitrary length paths in an RDF graph, potentially returning or matching the endpoints of the path. An IRI $p$ is a \emph{path expression}; if $e$, $e_1$ and $e_2$ are path expressions, then $\texttt{\^{}}e$ (inverse of $e$), $e_1 \texttt{/} e_2$ ($e_1$ followed by $e_2$), $e_1 \texttt{|} e_2$ ($e_1$ or $e_2$), $e\texttt{*}$ (zero or more $e$), $e\texttt{+}$ (one or more $e$), $e\texttt{?}$ (zero or one $e$), and $\texttt{(}$e$\texttt{)}$ (parentheses used for precedence) are also \textit{path expressions}; finally, if $p$, $p_1$, $\ldots$ $p_n$ are IRIs, then $\texttt{!}p$, $\texttt{!}\texttt{(}p_1 \texttt{|} \ldots \texttt{|} p_n\texttt{)}$ and $\texttt{!}\texttt{(}p_1 \texttt{|} \ldots \texttt{|} p_k \texttt{|}  \texttt{{\^{}}}p_{k+1} \texttt{|} \ldots \texttt{|} \texttt{{\^{}}}p_n\texttt{)}$ for $k+1 \leq n$ (negated property sets) are path expressions. Thereafter, a \textit{path pattern} $(s,e,o)$ where $e$ is a path expression is an expression. Other features supported in SPARQL~1.1 include sub-queries, negation, aggregation, value binding, and so forth; for brevity, we do not introduce definitions for all such features.

@@

We consider a dynamic RDF graph to be a sequence of $n$ versions of an RDF graph denoted $\mathcal{G} \da ( G_1, \ldots, G_n )$; in practice, we assume these versions to have regular intervals (e.g., hourly, daily, weekly, etc.). Further given a SPARQL (1.1) \texttt{SELECT} query $Q$, we would like to have knowledge of the dynamics of its results. The first such problem we consider -- which we call One Shot Change (OSC) -- accepts $\mathcal{G}$, $Q$ and a positive integer $k$ as input and outputs a boolean value predicting whether or not the results will change from $G_n$ to some future version $G_{n+k}$ (i.e., $Q(G_n) \neq Q(G_{n+k})$?). The second (more difficult) problem -- which we call Time-To-Live (TTL) -- accepts only $\mathcal{G}$ and $Q$ and outputs a positive integer $k$ as a prediction for the lowest such value where $Q(G_n) \neq Q(G_{n+k})$.

\ah{How do we define quality under bag semantics, without \texttt{DISTINCT}? Do we consider equality of unique results, or multiplicity of results?}

@@

\paragraph{Architecture:} We propose a method to predict answers for the OSC and TTL problems. Our method primarily considers static features of the query: these include statistics such as the number of triple patterns, as well as the presence of query operators, such as \texttt{UNION}, \texttt{FILTER}, specific path expressions, etc.; our model can also consider statistics about predicates in the dynamic RDF graph, as well as in which past versions of the dynamic graph the results for the query changed. Figure~\ref{fig:schema} shows the components of our approach, which receives as input a SPARQL query (the architecture itself already encodes information about the dynamic graph and the value of $k$ for OSC, which we set in experiments to 1, predicting changes in the next version). The Feature Extractor decomposes the query into a set of numeric features, described presently. The Dynamics Model accepts some features of the query, as well as information from the dynamic graph, in order to compose further features. The final features are passed to a classifier in the case of the OSC problem, and to a regression model in the case of the TTL problem, to generate the final output prediction. We now describe the Feature Extractor and Dynamics Model in more detail.

\begin{figure*}[!t]
	\centering
	\includegraphics[width=0.9\linewidth]{img/schema.pdf}
	\caption{Proposed architecture with Feature Extractor and Dynamics Model \ah{What is $g$ here and why is it output from the Dynamics Model (rather than input)?}}
	\label{fig:schema}
\end{figure*}

\paragraph{Feature Extractor:} Our initial set of features are based on analysis of the input SPARQL query. Figure~\ref{fig:query} shows the list of evaluated features, as well as -- for the purposes of illustration -- their values for an accompanying example query.

The first four features capture statistics about the query, indicating loosely its size and complexity. One may consider that the higher these four values are, in general, the more dynamic we can expect the query to be since there are more ``opportunities'' for the query results to be affected by change. In some cases, however, the hypothesized correlation is not direct since for example adding more triple patterns may serve to narrow the query down and focus it on a static part of the graph (e.g., when looking for the movies of directors, adding a triple pattern to restrict the results to directors who have died). Hence it will be of interest to see how these features affect the predictions made experimentally.

The next six features indicate the query operators used; to reduce the dimensionality of the problem, we capture only a subset of popularly-used features and group related features into one dimension: in the case of \textit{Recursive path}, we group queries that contain as expression of the form $e$\texttt{*} or $e$\texttt{+}. On the other hand, in \textit{Negation}, we group non-monotonic features that allow for modeling difference (\texttt{MINUS}, \texttt{NOT EXISTS}, \texttt{OPTIONAL}, \texttt{!BOUND}). These features -- though course-grained -- are straightforward to extract from a query, and offer valuable insights into how the query may behave in a dynamic query; for example, the \textit{Negation} feature captures information about the (non-)monotonicity of the query, while we suppose that \textit{Recursive paths}, which may traverse an arbitrary number of triples in the graph, might be sensitive to change. Again, such relations are not direct where experiments are required for further insights.

@@

@@

\paragraph{Results:} Table~\ref{tab:classifier} shows the $F_1$-measure for the predictions made; for reference, we include a baseline that randomly guesses yes/no. The results show that all configurations clearly outperform the random baseline; however, with the highest result being an $F_1$-measure of 0.706, we can also see that predicting changes in query results is indeed a challenging problem. Looking in further detail, as can be expected, prediction performs best when the \textsc{r} feature -- with knowledge of changes in the historical value of results of the input query -- is available. However, the \textsc{qp} features -- which do not assume the availability of such information -- are often competitive: more specifically, the best result for \textsc{r} is 0.706 with Linear SVM, while the best results for \textsc{qp} is 0.651 with Decision Trees.

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{enumitem}
\usepackage{gensymb}
\usepackage{amsmath}
\usepgfplotslibrary{clickable}
\usetikzlibrary{shapes,arrows,positioning,fit,backgrounds,matrix,chains,patterns,trees,calc,arrows.meta}

@@

\begin{document}
%
\title{Estimating the dynamics of\\SPARQL query results over Wikidata\thanks{This work was supported by CONICYT PFCHA, Doctorado Becas Chile/2017 -- 72180000, the Millennium Institute for Foundational Research on Data (IMFD) and FONDECYT Grant No. 1181896.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Alberto Moya \and
Aidan Hogan\orcidID{0000-0001-9482-1982}}
%
\authorrunning{A. Moya and A. Hogan}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{IMFD; DCC, University of Chile\\
\email{\{amoya,ahogan\}@dcc.uhile.cl}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
We address the problem of estimating when the results of an input SPARQL query results over dynamic RDF datasets will change. We propose a framework that extracts features from the query and/or from past versions of the target dataset. The resulting features are then used to predict (1) whether or not the results for a query will change in the near future using classifiers, and (2) when they might change (if at all) using linear regression. To evaluate our proposal, we create a gold standard based on 23 versions of Wikidata and a curated collection of 221 SPARQL queries. Our results show that the quality of predictions using query features and lightweight statistics of the dynamics of individual predicates is competitive with that obtained using (more costly to derive) knowledge of the complete historical changes in the query results.

@@

Some works \ah{Which? Cite at the end of the sentence} have proposed to predict dynamics based on the historical evolution of entities, but following such an approach for SPARQL queries is expensive because (1) a SPARQL query may involve potentially many entities; and (2) it is necessary to have the previous complete versions of data to analyse the entities relevant to a query. Recent works have explored the dynamics of Linked Data with the intention of finding patterns that allow for characterizing, recognizing, and predicting changes based on analysis of the domains, predicates, and schema~\cite{UmbrichHHPD10,KaferAUOH13,NishiokaS16,NishiokaS17,GonzalezH18}. Among these works, we can find hybrid approaches that have been developed to return fresher query results with faster response times by decomposing a query into dynamic sub-queries executed remotely, and static sub-queries executed over local caches~\cite{UmbrichKHP12}, but such an approach only focuses on estimating the dynamics of individual triple patterns, and does not consider, for example, query operators. More generally, coping with changes in remote data still presents a major challenge for applications leveraging dynamic Linked Data.

In this work, we address the problem of estimating if and/or when SPARQL query results will change in a future version of a dynamic RDF dataset. We propose a general architecture based on Machine Learning that extracts static features from a query, as well as dynamic features from the query and dataset. These features are fed into a classifier to predict whether or not the query results will change in a fixed point in the future, or, more ambitiously, can be fed into a regression model to predict when the query results are likely to change. With respect to the features used, we propose that there is a trade-off between those that are coarse-grained but easy to compute (e.g., static query features) versus those that are fine-grained but more costly to compute (e.g., historical changes in query results), with further features falling in the middle (e.g., high-level statistics about the dynamics of triples with particular predicates). Per this trade-off, which features to use for predicting the dynamics of query results may then depend on the particular application.

In order to better understand this conceptual trade-off -- and more generally, to evaluate the quality of predictions made by our framework -- we create a novel gold standard based on 23 weekly versions of the Wikidata knowledge graph~\cite{VrandecicK14} and 221 user-generated queries; we show this gold standard to have a variety of desirable features, including (most importantly) a balance of queries whose results never change, always change, as well as non-trivial queries whose results intermittently change. Using this gold standard, our experiments show that while the most accurate predictions are based on historical changes to the input query's results, predictions based only on query features and predicate dynamics remain competitive (while being much more efficient to maintain/extract).

\paragraph{Contributions:} (1) We propose a general architecture for predicting when/if the results of a SPARQL query will change in a future version of a dynamic RDF dataset. (2) We propose a number of features to instantiate this architecture based on analysis of the query, analysis of the dynamics of predicates in the data, and analysis of historical changes in the query results. (3) We create a gold standard for these tasks based on 23 weekly versions of Wikidata and a set of 221 real-world SPARQL queries. (4) We use this gold standard to compare the predictions obtained using different types of features and classifiers.

\paragraph{Paper structure:} In Section~\ref{sec:related}, we review related works. Section~\ref{sec:preliminar} provides a formal statement of the problem. Section~\ref{sec:approach} describes our proposed architecture and features. In Section~\ref{sec:data}, we describe the creation of our gold standard based on Wikidata. Section~\ref{sec:eval} presents the design and results of the experiments. We conclude and review future directions in Section~\ref{sec:conclusion}.

@@

Specifically regarding the dynamics of SPARQL query results, Passant and Mendes~\cite{PassantM10} proposed \textsc{sparqlPuSH} as a notification framework based  on \textsc{PubSubHubBub} (recently standardized as \textsc{WebSub}~\cite{webSub18}) aiming for strong consistency. On the other hand, aiming for weak consistency, Umbrich et al.~\cite{UmbrichKHP12,UmbrichKPPH12,ekawUmbrichKHP12} proposed various methods to obtain knowledge about dynamics for different query patterns, mainly based on predicates. Dehghanzadeh et al.~\cite{DehghanzadehPKUHD14} later proposed a method to estimate the freshness using cardinality estimation techniques based on predicates and characteristic sets. Combining the notion of subscription-based notifications and predicting dynamics, Knuth et al.~\cite{KnuthHS16} propose a middleware to which consumers may subscribe that periodically ranks and schedules refreshes for queries according to policies that take into account how likely the results are to be stale, how long ago the query was last refreshed, how many results previously changed, how long the query takes to run, etc.

\paragraph{Novelty:} Given a query and a dynamic RDF dataset, we aim to predict (1) whether or not the query's results will have changed in a fixed point in the near future and/or (2) when the results will change. Our work thus complements existing works aiming for weak consistency, but (i) generalizes the problem, proposing a framework that can incorporate statistics on predicate dynamics~\cite{UmbrichKHP12,UmbrichKPPH12,ekawUmbrichKHP12,DehghanzadehPKUHD14} and historical changes in query results~\cite{KnuthHS16}, as proposed in prior works, as well as novel types of features, (ii) introduces new features based on query operators and statistics; (iii) creates a novel gold standard based on Wikidata and presents comparative results that indicate the relative predictive power inherent in different types of features. Furthermore, to the best of our knowledge, no prior work uses Machine Learning models to predict changes in SPARQL query results. \ah{Is this last claim true? We should try to be as sure as possible.}

%investigates the potential of using static features of a query as a lightweight alternative with the intuition that, for example, queries with more triple patterns or certain query operators are more likely to have dynamic results. For the purposes of comparison, we complement these static features with further knowledge of the dynamics of predicates~\cite{UmbrichKHP12,UmbrichKPPH12,ekawUmbrichKHP12,DehghanzadehPKUHD14} and knowledge of how results for the input query have changed over past versions~\cite{KnuthHS16}, as considered in previous works. We feed these data into Machine Learning classification and regression models to predict future changes (hence, for the purposes of supervised learning, we require knowledge of the changes of some queries over time, though not necessarily for the input query). The benefits of such an approach can be seen in a number of ways: (i) as complementing existing approaches; (ii) as an alternative to predicate-based statistics, which requires analysis of prior versions of the underlying data; (iii) as an alternative to historical results, which requires having a record of (or computing at runtime) the results of the query over several historical versions. We further remark that, to the best of our knowledge, no existing work aims to predict the dynamics of SPARQL query results using Machine Learning models. \ah{I should make this more concise later.}

@@

SPARQL is the recommended query language to retrieve and manipulate data stored in the RDF format. In this work, we focus on SPARQL \texttt{SELECT} queries, where we will first define a SPARQL 1.0 query. Let $\Vars$ be a set of variables disjoint from the set of RDF terms. A \emph{SPARQL expression} is built recursively as follows. (1) A \emph{triple pattern} $t \in  (\Iris \cup \Bnodes \cup \Lits \cup \Vars) \times (\Iris \cup \Vars) \times (\Iris \cup \Lits \cup \Bnodes \cup \Vars)$ is an \emph{expression}. (2) If $Q_1$ and $Q_2$ are expressions and $R$ is a filter condition, then $Q_1$ \texttt{FILTER} $R$, $Q_1$ \texttt{UNION} $Q_2$, $Q_1$ \texttt{OPTIONAL} $Q_2$, $Q_1$ \texttt{AND} $Q_2$ are $expressions$. Finally, if $Q$ is an expression, $V$ a list of variables and $\Delta$ a boolean value, $\texttt{SELECT}_V^\Delta Q$ is a SPARQL \texttt{SELECT} query, where $V$ denotes the projected variables, and $\Delta$ the \texttt{DISTINCT} option that when true, removes duplicate results. The semantics of a SPARQL \texttt{SELECT} query $Q$ is defined in terms of its \emph{evaluation} over an RDF graph $G$, denoted $Q(G)$, giving a set of partial mappings from projected variables to the set $\Iris \cup \Lits \cup \Bnodes$; we refer to P{\'{e}}rez et al.~\cite{PerezAG09} for definitions. (According to the standard, a SPARQL query is evaluated over a set of \emph{named graphs}; however, since the Wikidata query service indexes data in one RDF graph, to keep discussion concise, we consider RDF graphs, though our methods can be generalized in a straightforward manner to the named graphs setting.)

Our method supports SPARQL 1.1 \texttt{SELECT} queries, which allow a variety of additional features. One key feature in this extension is that of \textit{property paths}, which allows for matching arbitrary length paths in an RDF graph, potentially returning or matching the endpoints of the path. An IRI $p$ is a \emph{path expression}; if $e$, $e_1$ and $e_2$ are path expressions, then $\texttt{\^{}}e$ (inverse of $e$), $e_1 \texttt{/} e_2$ ($e_1$ followed by $e_2$), $e_1 \texttt{|} e_2$ ($e_1$ or $e_2$), $e\texttt{*}$ (zero or more $e$), $e\texttt{+}$ (one or more $e$), $e\texttt{?}$ (zero or one $e$), and $\texttt{(}$e$\texttt{)}$ (parentheses used for precedence) are also \textit{path expressions}; finally, if $p$, $p_1$, $\ldots,$ $p_n$ are IRIs, then $\texttt{!}p$, $\texttt{!}\texttt{(}p_1 \texttt{|} \ldots \texttt{|} p_n\texttt{)}$ and $\texttt{!}\texttt{(}p_1 \texttt{|} \ldots \texttt{|} p_k \texttt{|}  \texttt{{\^{}}}p_{k+1} \texttt{|} \ldots \texttt{|} \texttt{{\^{}}}p_n\texttt{)}$ for $k+1 \leq n$ (negated property sets) are path expressions. Thereafter, a \textit{path pattern} $(s,e,o)$ where $e$ is a path expression is an expression. Other features supported in SPARQL~1.1 include sub-queries, negation, aggregation, value binding, and so forth; for brevity, we do not introduce definitions for all such features.

@@

We consider a dynamic RDF graph to be a sequence of $n$ versions of an RDF graph denoted $\mathcal{G} \da ( G_1, \ldots, G_n )$; in practice, we assume these versions to have regular intervals (e.g., hourly, daily, weekly, etc.). Further given a SPARQL (1.1) \texttt{SELECT} query $Q$, we would like to have knowledge of the dynamics of its results. The first such problem we consider -- which we call One Shot Change (OSC) -- accepts $\mathcal{G}$, $Q$ and a positive integer $k$ as input and outputs a boolean value predicting whether or not the results will change from $G_n$ to some future version $G_{n+k}$ (i.e., $Q(G_n) \neq Q(G_{n+k})$?). The second (more difficult) problem -- which we call Time-To-Live (TTL) -- accepts only $\mathcal{G}$ and $Q$ and outputs a positive integer $k$ as a prediction for the lowest such value where $Q(G_n) \neq Q(G_{n+k})$; in the case of bag semantics, we define ``$\neq$'' in terms of bag inequality, meaning that two bags of query results are different if the multiplicity of any result differs. Our problem is thus concerned with predicting if/when the results of a query will change, rather than predicting how  -- or to what extent -- they will change.

\ah{Is this how we define quality under bag semantics, without \texttt{DISTINCT}? Do we consider equality of unique results, or multiplicity of results?}

@@

\paragraph{Architecture:} In Figure~\ref{fig:schema}, we provide an overview of our proposed architecture for making predictions with respect to the OSC and TTL tasks. A SPARQL query $Q$ and a dynamic RDF graph $\mathcal{G}$ are given as input. The first component parses the query and generates static features which include statistics about the number of elements in the query (e.g., number of triple patterns, variables, etc.) as well as the presence of particular query operators (e.g., \texttt{UNION}, \texttt{FILTER}, recursive path expressions, etc.). The second component may take as input some features of the query, as well as information from the dynamic graph, and produce a set of further features; this component may compute, for example, statistics about the dynamicity of predicates used in the query, or statistics about changes in the query results over past versions. The final features are passed to a classifier in the case of the OSC problem, and to a regression model in the case of the TTL problem, to generate the final output prediction. We will now describe in more detail the query and data features considered in this work.

%\begin{figure*}[!t]
%	\centering
%	\includegraphics[width=0.9\linewidth]{img/schema.pdf}
%	\caption{Proposed architecture with Feature Extractor and Dynamics Model \ah{What is $g$ here and why is it output from the Dynamics Model (rather than input)?}}
%	\label{fig:schema}
%\end{figure*}

\begin{figure}[tb]
\newlength{\hgap}
\setlength{\hgap}{0.7cm}

\newcommand{\tb}[1]{\begin{tabular}{c} #1 \end{tabular}}
\newcommand{\hsp}{\vphantom{Ag}}

\tikzset{
    every node/.style={
    	font=\sffamily,
    },
	f/.style={ 
		draw,
		text centered,
		style={inner sep=0.8},
		font=\footnotesize,
		minimum width=1.3em,
		minimum height=1.2em
	},
	i/.style={ 
		draw,
		text centered,
		rounded corners
	},
	o/.style={ 
		draw,
		text centered,
		rounded corners
	},	
	c/.style={ 
		draw,
		top color=white, 
		bottom color=black!5
	},	
	arrout/.style={
		->,
		-latex,
		black!25},
	arrin/.style={
        <-,
        latex-,
		black!25},	
}

\centering
\resizebox{0.9\textwidth}{!}{
\begin{tikzpicture}
\node[i] (q) {$Q$};

\node[c,right=\hgap of q] (qfe) {\tb{Query \\ Feature \\ Extractor}}
  edge[arrin] (q);

\node[f,right=1.5\hgap of qfe] (f4) {\tiny ...};
\node[f,anchor=south] (f3) at (f4.north) {\tiny ...};
\node[f,anchor=south] (f2) at (f3.north) {\tiny ...};
\node[f,anchor=south] (f1) at (f2.north) {$f_1$};
\node[f,anchor=north] (f5) at (f4.south) {\tiny ...};
\node[f,anchor=north] (f6) at (f5.south) {\tiny ...};
\node[f,anchor=north] (fn) at (f6.south) {$f_n$};

\draw [arrout] (qfe.east) -- + (0.3,0) |- (f1);
\draw [arrout] (qfe.east) -- + (0.3,0) |- (f2);
\draw [arrout] (qfe.east) -- + (0.3,0) |- (f3);
\draw [arrout] (qfe.east) -- + (0.3,0) |- (f4);
\draw [arrout] (qfe.east) -- + (0.3,0) |- (f5);
\draw [arrout] (qfe.east) -- + (0.3,0) |- (f6);
\draw [arrout] (qfe.east) -- + (0.3,0) |- (fn);

\node[c,right=\hgap of fn] (dfe) {\tb{Data\\Feature\\Extractor}}
  edge[arrin] (fn);

\draw[arrout] (f6.east) -- (dfe.west|-f6.east);

\node[i,anchor=south] (g) at (q.mid|-dfe.south) {$\mathcal{G}$}
  edge[arrout] (dfe.west|-g.east);

\node[f,right=\hgap of dfe] (dfk) {$f'_k$}
  edge[arrin] (dfe);    
\node[f,anchor=south] (df6) at (dfk.north) {\tiny ...}
  edge[arrin] (dfe.east|-df6);  
\node[f,anchor=south] (df5) at (df6.north) {$f_j$}
  edge[arrin] (f5);    
\node[f,anchor=south] (df4) at (df5.north) {\tiny ...}
  edge[arrin] (f4);  
\node[f,anchor=south] (df3) at (df4.north) {\tiny ...}
  edge[arrin] (f3);      
\node[f,anchor=south] (df2) at (df3.north) {\tiny ...}
  edge[arrin] (f2);   
\node[f,anchor=south] (df1) at (df2.north) {$f_1$}
  edge[arrin] (f1);

\node[c,right=1.5\hgap of df2,minimum width=6.5em] (c) {Classification\hsp};

\draw [arrout] (df1.east) -- + (0.3,0) |- (c);
\draw [arrout] (df2.east) -- + (0.3,0) |- (c);
\draw [arrout] (df3.east) -- + (0.3,0) |- (c);
\draw [arrout] (df4.east) -- + (0.3,0) |- (c);
\draw [arrout] (df5.east) -- + (0.3,0) |- (c);
\draw [arrout] (df6.east) -- + (0.3,0) |- (c);
\draw [arrout] (dfk.east) -- + (0.3,0) |- (c);

\node[c,right=1.5\hgap of df6,minimum width=6.5em] (r) {Regression\hsp};

\draw [arrout] (df1.east) -- + (0.3,0) |- (r);
\draw [arrout] (df2.east) -- + (0.3,0) |- (r);
\draw [arrout] (df3.east) -- + (0.3,0) |- (r);
\draw [arrout] (df4.east) -- + (0.3,0) |- (r);
\draw [arrout] (df5.east) -- + (0.3,0) |- (r);
\draw [arrout] (df6.east) -- + (0.3,0) |- (r);
\draw [arrout] (dfk.east) -- + (0.3,0) |- (r);

\node[o,right=\hgap of c] (osc) {\tb{OSC \\ Prediction}}
  edge[arrin] (c);

\node[o,right=\hgap of r] (ttl) {\tb{TTL \\ Prediction}}
  edge[arrin] (r);

\end{tikzpicture}
}

\caption{Proposed architecture for predicting change in query results\label{fig:schema}}
\end{figure}

\paragraph{Query Features:} Our initial set of features are based on analysis of the input SPARQL query. Figure~\ref{fig:query} shows the list of evaluated features, as well as -- for the purposes of illustration -- their values for an accompanying example query.

The first four features capture statistics about the query, indicating loosely its size and complexity. One may consider that the higher these four values are, in general, the more dynamic we can expect the query to be since there are more ``opportunities'' for the query results to be affected by change. In some cases, however, the hypothesized correlation is not direct since, for example, adding more triple patterns may serve to narrow the query down and focus it on a static part of the graph (e.g., when looking for the movies of directors, adding a triple pattern to restrict the results to directors who have died). Hence it will be of interest to see how these features affect the predictions made experimentally.

The next six features indicate the query operators used; to reduce the dimensionality of the problem, we capture only a subset of popularly-used features and group related features into one dimension: in the case of \textit{Recursive path}, we group queries that contain as expression of the form $e$\texttt{*} or $e$\texttt{+}. On the other hand, in \textit{Negation}, we group non-monotonic features that allow for modeling difference (\texttt{MINUS}, \texttt{NOT EXISTS}, \texttt{OPTIONAL}, \texttt{!BOUND}). These features -- though course-grained -- are straightforward to extract from a query, and offer valuable insights into how the query may behave in a dynamic query; for example, the \textit{Negation} feature captures information about the (non-)monotonicity of the query, while we suppose that \textit{Recursive paths}, which may traverse an arbitrary number of triples in the graph, might be more sensitive to change. Again, such correlations are not without exception and will require experimental scrutiny.

@@

\paragraph{Data Features (result dynamics):} The second data feature we capture indicates the past versions of the dynamic graph in which the query results $Q$ changed (what Knuth et al. call the ``clairvoyant model''~\cite{KnuthHS16}). \ah{How precisely is this represented?} While this feature offers rich information for prediction, given a query that has not previously been seen, it may be costly to compute, potentially requiring the execution of the query over each version, which in turn requires maintaining indexes over a variety of past versions.

@@

\paragraph{Results:} Table~\ref{tab:classifier} shows the $F_1$-measure for the predictions made; for reference, we include a baseline that randomly guesses yes/no. With the highest result being an $F_1$-measure of 0.706, the results show that predicting changes in query results is indeed a challenging problem; on the other hand, all configurations clearly outperform the random baseline. Looking in further detail, as can be expected, prediction performs best when the \textsc{r} feature -- with knowledge of changes in the historical value of results of the input query -- is available. However, the \textsc{qp} features -- which do not assume the availability of such information -- are often competitive: more specifically, the best result for \textsc{r} is 0.706 with Linear SVM, while the best results for \textsc{qp} is 0.651 with Decision Trees.
